---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r load-and-preprocess}
# Load the Data
data <- read.csv("./activity.csv")
head(data)

# Inspect the Data Structure
str(data)
summary(data)

# Check for Missing Values
missing_steps <- sum(is.na(data$steps))
print(paste("Total missing values in 'steps':", missing_steps))

# Data Transformation
data$date <- as.Date(data$date, format="%Y-%m-%d")

# Initial Exploration of Missing Values
table_missing_dates <- table(is.na(data$steps), data$date)
print("Table of missing values by date:")

# Data Integrity Check
boxplot(data$steps ~ data$date, main="Boxplot of Steps by Date")
invalid_intervals <- any(data$interval < 0 | data$interval > 2355)
print(paste("Are there any invalid intervals? ", invalid_intervals))

## What is mean total number of steps taken per day?

```{r mean-total-steps, echo=TRUE}
library(dplyr)

# Group data by date and calculate total steps per day
daily_steps <- data %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps, na.rm = TRUE))

# Histogram of total steps per day
hist(daily_steps$total_steps, 
     main = "Histogram of Total Steps Per Day", 
     xlab = "Total Steps", 
     col = "blue")

# Calculate and report the mean and median
mean_steps <- mean(daily_steps$total_steps)
median_steps <- median(daily_steps$total_steps)

print(paste("Mean of total steps per day:", mean_steps))
print(paste("Median of total steps per day:", median_steps))

## What is the average daily activity pattern?

```{r average-daily-pattern, echo=TRUE}
library(dplyr)
library(ggplot2)

# Calculate average steps for each interval
average_interval_steps <- data %>%
  group_by(interval) %>%
  summarize(average_steps = mean(steps, na.rm = TRUE))

# Time Series Plot of the Average Number of Steps per Interval
ggplot(average_interval_steps, aes(x = interval, y = average_steps)) +
  geom_line(color = "blue") +
  labs(title = "Average Daily Steps per 5-minute Interval",
       x = "Interval",
       y = "Average Number of Steps") +
  theme_minimal()

# Identify the interval with the maximum average steps
max_interval <- average_interval_steps[which.max(average_interval_steps$average_steps), ]
print(paste("The 5-minute interval with the maximum average number of steps is:", max_interval$interval))

## Imputing missing values

```{r impute-missing-values, echo=TRUE}
library(dplyr)

# Total number of missing values
total_missing_values <- sum(is.na(data$steps))
print(paste("Total number of missing values in the dataset:", total_missing_values))

# Strategy for filling missing values: Use the mean for 5-minute interval
imputed_data <- data
for (i in unique(data$interval)) {
  imputed_data$steps[is.na(imputed_data$steps) & imputed_data$interval == i] <- 
    mean(data$steps[data$interval == i], na.rm = TRUE)
}

# Checking if all missing values are filled
total_missing_values_after <- sum(is.na(imputed_data$steps))
print(paste("Total number of missing values after imputation:", total_missing_values_after))

# Histogram of the total number of steps taken each day with imputed data
daily_steps_imputed <- imputed_data %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps))

hist(daily_steps_imputed$total_steps, 
     main = "Histogram of Total Steps Per Day (With Imputed Data)", 
     xlab = "Total Steps", 
     col = "green")

# Calculate and report the mean and median with imputed data
mean_steps_imputed <- mean(daily_steps_imputed$total_steps)
median_steps_imputed <- median(daily_steps_imputed$total_steps)

print(paste("Mean of total steps per day (with imputed data):", mean_steps_imputed))
print(paste("Median of total steps per day (with imputed data):", median_steps_imputed))

# Compare with original estimates
print("Original estimates without imputation:")
print(paste("Mean:", mean_steps))
print(paste("Median:", median_steps))

## Are there differences in activity patterns between weekdays and weekends?

```{r weekday-weekend-patterns, echo=TRUE}
library(dplyr)
library(ggplot2)
library(lubridate)

# Create a factor variable for weekday or weekend
imputed_data$day_type <- ifelse(weekdays(imputed_data$date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
imputed_data$day_type <- as.factor(imputed_data$day_type)

# Calculate average steps for each interval by day type
average_interval_steps_day_type <- imputed_data %>%
  group_by(interval, day_type) %>%
  summarize(average_steps = mean(steps))

# Panel plot showing the time series plot of average steps by day type
ggplot(average_interval_steps_day_type, aes(x = interval, y = average_steps, color = day_type)) +
  geom_line() +
  facet_wrap(~ day_type, ncol = 1, scales = "free_y") +
  labs(title = "Average Number of Steps per 5-minute Interval by Day Type",
       x = "Interval",
       y = "Average Number of Steps") +
  theme_minimal() +
  theme(legend.position = "bottom")